\documentclass[12pt]{book}
\begin{document}
\title{Triumph of the Church\\(History of Heresies and its Refutation)}
\author{St. Alphonsus Marie de Liguori}
\date{}
\maketitle
\noindent{
Dublin: Published by James Duffy, 10, Wellington-Quay. 1847.
\\Printed by William Holdin, 10, Abbey-street.}

\chapter*{Translator's Preface}
THE ardent wish manifested by "the Faithful for an acquaintance with the valuable writings of ST. 
LIGUORI, induced me to undertake the Translation of his History of Heresies, one of his greatest works.
The Holy Author was induced to write this Work, to meet the numbers of infidel publications, with
which Europe was deluged in the latter half of the last century. Men’s minds were then totally unsettled;
dazzled by the glare of a false philosophy, they turned away from the light of the Gospel. The heart of the
Saint was filled with sorrow, and he laboured to avert the scourge he saw impending over the unfaithful
people. He implored the Ministers of his Sovereign to put the laws in force, preventing the introduction
of irreligious publications into the Kingdom of Naples, and he published this Work, among* others, to
prove, as he says, that the Holy Catholic Church is the only true one the Mistress of Truth the Church,
founded by Jesus Christ himself, which would last to the end of time, notwithstanding the persecutions
of the infidel, and the rebellion of her own heretical children. He dedicates the Book to the Marquis
Tanucci, the Prime Minister of the Kingdom, whom he praises for his zeal for Religion, and his vigorous
execution of the laws against the vendors of infidel publications. He brings down the History from the
days of the Apostles to his own time, concluding- with the Refutation of the Heresies of Father Berruyer.

I have added a Supplementary Chapter, giving- a succinct account of the Heretics and Fanatics of the last
eighty years. It was, at first, my intention to make it more diffuse; but, then, I considered that it would be
out of proportion with the remainder of the Work. This Book may be safely consulted, as a work of
reference: the Author constantly quotes his authorities; and the Student of Ecclesiastical History can at
once compare his statements with the sources from which he draws. In the latter portion of the Work, and
especially in that portion of it, the most interesting- to us, the History of the English Reformation, the
Student may perceive some slight variations between the original text and my translation. I have collated
the Work with the writings of modern Historians the English portion, especially with Hume and Lingard
and wherever I have seen the statements of the Holy Author not borne out by the authority of our own
Historians, I have considered it more prudent to state the facts, as they really took place; for our own
writers must naturally be supposed to be better acquainted with our History, than the foreign authorities
quoted by the Saint. The reader will also find the circumstances, and the names of the actors, when I
considered it necessary, frequently given more in detail than in the original.

In the style, I have endeavoured, as closely as the genius of our language would allow, to keep to the
original. St. Alphonsus never sought for ornament; a clear, lucid statement of facts is what he aimed at;
there is nothing inflated in his writings; he wrote for the people, and that is the principal reason, I
imagine, why not only his Devotional Works, but his Historical and Theological Writings, also, have been
in such request: but, while he wrote for the people, we are not to imagine that he did not also please the
learned. His mind was richly stored with various knowledge; he was one of the first Jurists of his day; his
Theological science elicited the express approbation of the greatest Theologian of his age Benedict XIV.;
he was not only a perfect master of his own beautiful language, but profoundly read in both Greek and
Latin literature also, and a long life constantly employed in studies, chiefly ecclesiastical, qualified him,
above any man of his time, to become an Ecclesiastical Historian, which no one should attempt unless he
be a general I might almost say a universal, scholar : so much for the Historical portion of the Work.

In the Second Part, the Refutation of Heresies, the Holy Author comprises, in a small space, a vast
amount of Theological information; in fact, there is no Heresy which cannot be refuted from it. Not alone
are the usual Heresies, which we have daily to combat such as those opposed to the Real Presence, the
Authority of the Church, the doctrine of Justification, clearly and diffusely refuted, but those abstruse
heretical opinions concerning- Grace, Free Will, the Procession of the Holy Ghost, the Mystery of the
Incarnation, and the two Natures of Christ, and so forth, are also clearly and copiously confuted; the
intricacies of Pelagianism, Calvinism, and Jansenism, are unravelled, and the true Doctrine of the Church
triumphantly vindicated. The reader will find, in general, the quotations from the Fathers in the original,
but those unacquainted with Latin will easily learn their sentiments from the text. The Scripture 
quotations are from the Douay Version.

Every Theologian will be aware of the difficulty of giving- scholastic terms in an English dress. In the
language of the Schools, the most abstract ideas, which would require a sentence to explain them in our
tongue, are most appropriately expressed by a single word; all the Romance languages, daughters of the
Latin, have very nearly the same facility, but our Northern tongue has not, I imagine, flexibility enough
for the purpose. I have, however, endeavoured, as far as I could, to preserve the very terms of the
original, knowing how easy it is to give a heterodox sense to a passage, by even the most trivial deviation
from the very expression of the writer. The Theological Student will thus, I hope, find the Work a
compact Manual of Polemic Theology; the Catholic who, while he firmly believes all that the Church
teaches, wishes to be able to give an account of the Faith that is in him, will here find it explained and
defended; while those not of the " fold," but for whom we ardently pray, that they may hear the voice of
the " one Shepherd," may see, by its attentive perusal, that they inhabit a house " built upon the sand,"
and not the house " on the rock."

They will behold the mighty tree of Faith sprung from the grain of mustard-seed planted by our
Redeemer, always flourishing always extending*, neither uprooted by the storms of persecution, nor
withered by the sun of worldly prosperity. Nay more, the very persecution the Church of God has
suffered, and is daily enduring, only extends it more and more; the Faithful, persecuted in " one city," fly
elsewhere, bearing with them the treasure of Faith, and communicating it to those among- whom they
settle, as the seeds of fertility are frequently borne on the wings of the tempest to the remote desert,
which would otherwise be cursed with perpetual barrenness. The persecution of the Church in Ireland,
for example, "has turned the desert into fruitfulness," in America, in Australia, in England itself, and the
grey mouldering ruins of our fanes on the hill sides are compensated for by the Cathedral Churches
across the ocean. The reader will see Heresy in every age, from the days of the Apostles themselves down
to our own time, rising up, and vanishing after a while, but the Church of God is always the same, her
Chief Pastors speaking with the same authority, and teaching the same doctrine to the trembling
Neophytes’ in the Catacombs, and to the C{\ae}sars on the throne of the world. Empires are broken into
fragments and perish nations die away, and are only known to the historian languages spoken by
millions disappear every thing that is man’s work dies like man; heresies, like the rest, have their rise,
their progress, their decay, but Faith alone is eternal and unchangeable, "yesterday, to-day, and the same
for ever."

\chapter*{Author's Preface}
\begin{enumerate}
\item My object in writing this work is to prove that the Roman Catholic Church is the only true one among
so many other Churches, and to show how carefully the Almighty guarded her, and brought her
victoriously through all the persecutions of her enemies. Hence, as St. Ir{\ae}neus says (Lib. 3, cap. 3, n. 2), all
should depend on the Roman Church as on their fountain and head. This is the Church founded by Jesus
Christ, and propagated by the Apostles; and although in the commencement persecuted and contradicted
by all, as the Jews said to St. Paul in Rome : " For as concerning this sect (thus they called the Church), we
know that it is gainsayed every where" (Acts, xxviii, 22); still she always remained firm, not like the other
false Churches, which in the beginning numbered many followers, but perished in the end, as we shall
see in the course of this history, when we speak of the Arians, Nestorians, Eutychians, and Pelagians; and
if any sect still reckons many followers, as the Mahometans, Lutherans, or Calvinists, it is easy to see that
they are upheld, not by the love of truth, but either by popular ignorance, or relaxation of morals. St.
Augustine says that heresies are only embraced by those who had they persevered in the faith, would be
lost by the irregularity of their lives (St. Aug. de Va. Rel. c. 8.)

\item Our Church, on the contrary, notwithstanding that she teaches her children a law opposed to the
corrupt inclinations of human nature, not only never failed in the midst of persecutions, but even gained
strength from them; as Tertullian (Apol. cap. ult.) says, the blood of martyrs is the seed of Christians, and
the more we are mown down the more numerous we become; and in the 20th chapter of the same work
he says, the kingdom of Christ and his reign is believed and he is worshipped by all nations. Pliny the
Younger confirms this in his celebrated Letter to Trajan, in which he says that in Asia the temples of the
gods were deserted, because the Christian Religion had overrun not only the cities but even the villages.

\item This, certainly, never could have taken place without the power of the Almighty, who intended to
establish in the midst of idolatry, a new religion, to destroy all the superstitions of the false religion, and
the ancient belief in a multitude of false gods adored by the Gentiles, by their ancestors, by the
magistrates, and by the emperors themselves, who made use of all their power to protect it, and still the
Christian faith was embraced by many nations who forsook a relaxed law for a hard and difficult one,
forbidding them to pamper their sensual appetites. What but the power of God could accomplish this?

\item Great as the persecutions were which the Church suffered from idolatry, still greater were those she
had to endure from the heretics which sprang from her own bosom, by means of wicked men, who, either
through pride or ambition, or the desire of sensual license, endeavoured to rend the bowels of their
parent. Heresy has been called a canker : " It spreadeth like a canker" (II. Tim. ii, 17); for as a canker
infects the whole body, so heresy infects the whole soul, the mind, the heart, the intellect, and the will. It
is also called a plague, for it not only infects the person contaminated with it, but those who associate
with him, and the fact is, that the spread of this plague in the world has injured the Church more than
idolatry, and this good mother has suffered more from her own children than from her enemies. Still she
has never perished in any of the tempests which the heretics raised against her; she appeared about to
perish at one time through the heresy of Arius, when the faith of the Council of Nice, through the
intrigues of the wicked Bishops, Valens and Ursacius, was condemned, and, as St. Jerome says, the world
groaned at finding itself Arian (1); and the Eastern Church appeared in the same danger during the time
of the heresies of Nestorius and Eutyches. But it is wonderful, and at the same time consoling, to read the
end of all those heresies, and behold the bark of the Church, which appeared completely wrecked and
sunk through the force of those persecutions, in a little while floating more gloriously and triumphantly
than before.

\item St. Paul says: " There must be heresies, that they also who are reproved may be made manifest among
you" (I. Cor. ii, 19). St. Augustine, explaining this text, says that as fire is necessary to purify silver, and
separate it from the dross, so heresies are necessary to prove the good Christians among the bad, and to
separate the true from the false doctrine. The pride of the heretics makes them presume that they know
the true faith, and that the Catholic Church is in error, but here is the mistake : our reason is not sufficient
to tell us the true faith, since the truths of Divine Faith are above reason; we should, therefore, hold by
that faith which God has revealed to his Church, and which the Church teaches, which is, as the Apostle
says, " the pillar and the ground of truth" (I. Tim. iii, 15).\\
\textbf{(1) St. Hieron. Dial, adversus Lucifer.}\\

Hence, as St. Ir{\ae}neus says, "It is necessary that all should depend on the Roman Church as their head and
fountain; all Churches should agree with this Church on account of her priority of principality, for there
the traditions delivered by the Apostles have always been preserved" (St. Iran, lib. 3, c. 3); and by the
tradition derived from the Apostles which the Church founded at Rome preserves, and the Faith
preserved by the succession of the Bishops, we confound those who through blindness or an evil
conscience draw false conclusions (Ibid). " Do you wish to know," says St. Augustine, " which is the true
Church of Christ? Count those priests who, in a regular succession have succeeded St. Peter, who is the
Rock, against which the gates of hell will not prevail" (St. Aug. in Ps. contra part Donat.) : and the holy
Doctor alleges as one of the reasons which detain him in the Catholic Church, the succession of Bishops to
the present time in the See of St. Peter" (Epis. fund, c. 4, n. 5); for in truth the uninterrupted succession
from the Apostles and disciples is characteristic of the Catholic Church, and of no other.

\item It was the will of the Almighty that the Church in which the true faith was preserved should be one,
that all the faithful might profess the one faith, but the devil, St. Cyprian says (2), invented heresies to
destroy faith, and divide unity. The enemy has caused mankind to establish many different churches, so
that each, following the faith of his own particular one, in opposition to that of others, the true faith might
be confused, and as many false faiths formed as there are different churches, or rather different
individuals. This is especially the case in England, where we see as many religions as families, and even
families themselves divided in faith, each individual following his own. St. Cyprian, then, justly says that
God has disposed that the true faith should be preserved in the Roman Church alone, so that there being
but one Church there should be but one faith and one doctrine for all the faithful. St. Optatus
Milevitanus, writing to Parmenianus, says, also : " You cannot be ignorant that the Episcopal Chair of St.
Peter was first placed in the city of Rome, in which one chair unity is observed by all" (St. Opt. I 2, cont.
Parmen.)\\
\textbf{(2) St. Cyprian de Unitate Ecclesi{\ae}.}

\item The heretics, too, boast of the unity of their Churches, but St. Augustine says that it is unity against
unity. " What unity," says the Saint, " can all those churches have which are divided from the Catholic
Church, which is the only true one; they are but as so many useless branches cut off from the Vine, the
Catholic Church, which is always firmly rooted. This is the One Holy, True, and Catholic Church,
opposing all heresies; it may be opposed, but cannot be conquered. All heresies come forth from it, like
useless shoots cut off from the vine, but it still remains firmly rooted in charity, and the gates of hell shall
not prevail against it" (St. Aug. lib. 1, de Symbol ad Cath. c. 6). St. Jerome says that the very fact of the
heretics forming a church apart from the Roman Church, is a proof, of itself, that they are followers of
error, and disciples of the devil, described by the Apostle, as " giving heed to spirits of error and doctrines
of devils" (I. Tim. iv, 1).

\item The Lutherans and Calvinists say, just as the Donatists did before them, that the Catholic Church
preserved the true faith down to a certain period some say to the third, some to the fourth, some to the
fifth century but that after that the true doctrine was corrupted, and the spouse of Christ became an
adulteress. This supposition, however, refutes itself; for, granting that the Roman Catholic Church was
the Church first founded by Christ, it could never fail, for our Saviour himself promised that the gates of
hell never should prevail against it : "I say unto you that you are Peter, and on this Rock I will build my
church, and the gates of hell shall not prevail against it" (Matt, xviii, 18). It being certain, then, that the
Roman Catholic Church was the true one, as Gerard, one of the first ministers of Luther, admits (Gerard
de Eccles. cap. 11, sec. 6) it to have been for the first five hundred years, and to have preserved the
Apostolic doctrine during that period, it follows that it must always have remained so, for the spouse of
Christ as St. Cyprian says, could never become an adulteress.

\item  The heretics, however, who, instead of learning from the Church the dogmas they should believe, wish
to teach her false and perverse dogmas of their own, say that they have the Scriptures on their side, which
are the fountain of truth, not considering, as a learned author (3) justly remarks, that it is not by reading,
but by understanding, them, that the truth can be found. Heretics of every sort avail themselves of the
Scriptures to prove their errors, but we should not interpret the Scripture according to our own private
opinions, which frequently lead us astray, but according to the teaching of the Holy Church which is
appointed the Mistress of true doctrine, and to whom God has manifested the true sense of the Divine
books. This is the Church, as the Apostle tells us, which has been appointed the pillar and the ground of
truth: \textit{\textbf{"that thou mayest know how thou oughtest to behave thyself in the house of God, which is the
Church of the living God, the pillar and the ground of truth"}} (I. Tim. iii, 15.) Hence St. Leo says that the
Catholic faith despises the errors of heretics barking against the Church, who deceived by the vanity of
worldly wisdom, have departed from the truth of the Gospel (St. Leo, Ser. 8, de Nat Dim.)

\item I think the History of Heresies is a most useful study, for it shows the truth of our Faith more pure
and resplendent, by showing how it has never changed; and if, at all times, this is useful, it must be
particularly so at present, when the most holy maxims and the principal dogmas of Religion are put in
doubt : it shows, besides, the care God always took to sustain the Church in the midst of the tempests
which were unceasingly raised against it, and the admirable manner in which all the enemies who
attacked it were confounded. The History of Heresies is also useful to preserve in us the spirit of humility
and subjection to the Church, and to make us grateful to God for giving us the grace of being born in
Christian countries; and it shows how the most learned men have fallen into the most grievous errors, by
not subjecting themselves to the Church’s teaching.\\
\textbf{(3) Danes, Gen. Temp. Nat. in Epil.}

\item I will now state my reasons for writing this Work; some may think this labour of mine superfluous,
especially as so many learned authors have written expressly and extensively the history of various
heresies, as Tertullian, St. Ir{\ae}neus, St. Epiphanius, St. Augustine, St. Vincent of Lerins, Socrates,
Sozymen, St. Philastrius, Theodoret, Nicephorus, and many others, both in ancient and modern times.
This, however, is the very reason which prompted me to write this Work; for as so many authors have
written, and so extensively, and as it is impossible for many persons either to procure so many and such 
expensive works, or to find time to read them, if they had them, I, therefore, judged it better to collect in a
small compass the commencement and the progress of all heresies, so that in a little time, and at little
expense, any one may have a sufficient knowledge of the heresies and schisms which infected the
Church. I have said in a small compass, but still, not with such brevity as some others have done, who
barely give an outline of the facts, and leave the reader dissatisfied, and ignorant of many of the most
important circumstances. I, therefore, have studied brevity; but I wish, at the same time, that my readers
may be fully informed of every notable fact connected with the rise and progress of, at all events, the
principal heresies that disturbed the Church.

\item Another reason I had for publishing this Work was, that as modern authors, who have paid most
attention to historical facts, have spoken of heresies only as a component part of Ecclesiastical History, as
Baronius, Fleury, Noel Alexander, Tillemont, Orsi, Spondanus, Raynaldus, Graveson, and others, and so
have spoken of each heresy chronologically, either in its beginning, progress, or decay, and, therefore, the
reader must turn over to different parts of the works to find out the rise, progress, and disappearance of
each heresy; I, on the contrary, give all at once the facts connected with each heresy in particular.

\item Besides, these writers have not given the Refutation of Heresies, and I give this in the second part of
the Work; I do not mean the refutation of every heresy, but only of the principal ones, as those of
Sabellius, Arius, Pelagius, Macedonius, Nestorius, Eutyches, the Monothelites, the Iconoclasts, the
Greeks, and the like. I will merely speak of the authors of other heresies of less note, and their falsity will
be apparent, either from their evident weakness, or from the proofs I bring forward against the more
celebrated heresies I have mentioned.

\item We ought, then, dear reader, unceasingly to thank our Lord for giving us the grace of being born and
brought up in the bosom of the Catholic Church. St. Francis de Sales exclaims: \textit{\textbf{ "O good God! many and
great are the benefits thou hast heaped on me, and I thank thee for them; but how shall I be ever able to
thank thee for enlightening me with thy holy Faith?"}} And writing to one of his friends, he says: \textit{\textbf{
"God! the beauty of thy holy Faith appears to me so enchanting, that I am dying with love of it, and I imagine I
ought to enshrine this precious gift in a heart all perfumed with devotion."}} St. Teresa never ceased to
thank God for having made her a daughter of the Holy Church: her consolation at the hour of death was
to cry out: \textit{\textbf{"I die a child of the Holy Church! I die a child of the Holy Church."}} We, likewise, should
never cease praising Jesus Christ for this grace bestowed on us one of the greatest conferred on us one
distinguishing us from so many millions of mankind, who are born and die among infidels and heretics :
\textit{\textbf{"He has not done in like manner to every nation"}} (Psalm cxlvii, 9). With our minds filled with gratitude
for so great a favour, we shall now see the triumph the Church has obtained through so many ages, over
so many heresies opposed to her. I wish to remark, however, before I begin, that I have written this Work
amidst the cares of my Bishoprick, so that I could not give a critical examination, many times, to the facts
I state, and, in such case, I give the various opinions of different authors, without deciding myself on one
side or the other. I have endeavoured, however, to collect all that could be found in the most correct and
notable writers on the subject; but it is not impossible that some learned persons may be better
acquainted with some facts than I am.
\end{enumerate}

\chapter{Heresies of the First Century}
\section{Simon Magus}
Simon Magus (1), the first heretic who disturbed the Church, was born in a part of Samaria called
Githon or Gitthis. He was called Magus, or the Magician, because he made use of spells to deceive the
multitude; and hence he acquired among his countrymen the extraordinary name of " The Great Power
of God" (Acts, viii, 1 0). " This man is the power of God which is called great." Seeing that those on whom
the Apostles Peter and John laid hands received the Holy Ghost, he offered them money to give to him
the power of communicating the Holy Ghost in like manner; and on that account the detestable crime of
selling holy things is called Simony. He went to Rome, and there was a statue erected to him in that city, a
fact which St. Justin, in his first Apology, flings in the face of the Romans : " In your royal city," he says, "
he (Simon) was esteemed a God, and a statue was erected to him in the Island of the Tyber, between the
two bridges, bearing this Latin inscription SIMONI, DEO SANCTO."\\
(1) Baron. Annal, 35, d. 23; N. Alex. Hist. Ecclesias. t. 5, c. 11, n.-l; Hermant. His. Con. 56, 1, c. 7; Van
Ranst, His. Her. n. 1.

Samuel Basnage, Petavius, Valesius, and many others, deny this fact; but Tillemont, Grotius, Fleury, and
Cardinal Orsi defend it, and adduce in favour of it the authority of Tertullian, St. Irenæus, St. Cyril of
Jerusalem, St. Augustine, Eusebius, and Theodoret, who even says the statue was a bronze one. Simon
broached many errors, which Noel Alexander enumerates and refutes (2). The principal ones were that
the world was created by angels; that when the soul leaves the body it enters into another body, which, if
true, says St. Iræneus (3), it would recollect all that happened when it inhabited the former body, for
memory, being a spiritual quality, it could not be separated from the soul. Another of his errors was one
which has been brought to light by the heretics of our own days, that man had no free will, and,
consequently, that good works are not necessary for salvation. Baronius and Fleury relate (4), that, by
force of magic spells, he one day caused the devil to elevate him in the air; but St. Peter and St. Paul being
present, and invoking the name of Jesus Christ, he fell down and broke both his legs. He was carried
away by his friends; but his corporeal and mental sufferings preyed so much on him, that, in despair, he
cast himself out of a high window; and thus perished the first heretic who ever disturbed the Church of
Christ (5). Basnage, who endeavours to prove that St. Peter never was in Rome, and never filled the
pontifical chair of that city, says that this is all a fabrication; but we have the testimony of St. Ambrose, St.
Isidore of Pelusium, St. Augustine, St. Maximus, St. Philastrius, St. Cyril of Jerusalem, Severus Sulpicius,
Theodoret, and many others, in our favour. We have, besides, a passage in Seutonius, which corroborates
their testimony, for he says (lib. VI., cap. xii), that, while Nero assisted at the public sports, a man
endeavoured to fly, but, after elevating himself for a while, he fell down, and the Emperors pavilion was
sprinkled with his blood.\\
(2) Nat. Alex. t. 5, in fin. Dis. 24.\\
(3) St. Iræneus, de Heresi. l. 2, c. 58.\\
(4) Baron. Ann. 35, n. 14, ad. 17; Fleury, His. Eccl. t. 1, l. 2, n. 23; St. Augus.; St. Joan. Chris.\\
(5) Baron, n. 17; Nat. Alex. t. 5, c. 11; Orsi, Istor. Eccl. l. 1, n. 20, and l. 2, n. 19; Berti. Brev. Histor. t. 1, c. 3.
\section{Menander}
Menander was a Samaritan likewise, and a disciple of Simon Magus; he made his appearance in the
year of our Lord 73. He announced himself a messenger from the " Unknown Power," for the salvation of
mankind. No one, according to him, could be saved, unless he was baptized in his name, and his baptism,
he said, was the true resurrection, so that his disciples would enjoy immortality even in this life (6).
Cardinal Orsi adds, that Menander was the first who invented the doctrine of "Eons," and that he taught
that Jesus Christ exercised human functions in appearance alone.
\section{Cerinthus}
Cerinthus was the next after Menander, but he began to broach his doctrine in the same year (7). His
errors can be reduced to four heads : he denied that God was the creator of the world; he asserted that the
law of Moses was necessary for salvation; he also taught that after the resurrection Jesus Christ would
establish a terrestrial kingdom in Jerusalem, where the just would spend a thousand years in the
enjoyment of every sensual pleasure; and, finally, he denied the divinity of Jesus Christ. The account
Bernini gives of his death is singular (8). The Apostle St. John, he says, met him going into a bath, when,
turning to those along with him, he said, let us hasten out of this, lest we be buried alive, and they had
scarcely gone outside when the whole building fell with a sudden crash, and the unfortunate Cerinthus
was overwhelmed in the ruins. One of the impious doctrines of this heretic was, that Jesus was a mere
man, born as all other men are, and that, when he was baptized in the river Jordan, Christ descended on
him, that is, a virtue or power, in form of a dove, or a spirit sent by God to fill him with knowledge, and
communicate it to mankind; but after Jesus had fulfilled his mission, by instructing mankind and
working miracles, he was deserted by Christ, who returned to heaven, and left him to darkness and
death. Alas ! what impiety men fall into when they desert the light of faith, and follow their own weak
imaginations.\\
(6) Fleury, loc. cit. n. 42; N. Alex. loc. cit. art. 2\\
(7) N. Alex. t. 5, c. 11, or. 5; Fleury, t. 1, L 2, n. 42; Berti, loc. cit. : Orsi, t. 1, l. 2, n. 43.\\
(8) Bernin. Istor. del Eresia, t. 1, c. 1; St. Iren. 1. 3, c. 4, de S.
\section{Ebion}
Ebion prided himself in being a disciple of St. Peter, and could even bear to hear St. Pauls name
mentioned. He admitted the sacrament of baptism; but in the consecration of the Eucharist he used
nothing but water in the chalice; he, however, consecrated the host in unleavened bread, and Eusebius
says he performed this every Sunday. According to St. Jerome, the baptism of the Ebionites was admitted
by the Catholics. He endeavoured to unite the Mosaic and Christian law, and admitted no part of the 
New Testament, unless the Gospel of St. Matthew, and even that mutilated, as he left out two chapters,
and altered the others in many places. The ancient writers say that St. John wrote his Gospel to refute the
errors of Ebion. The most impious of his blasphemies was, that Jesus Christ was the son of Joseph and
Mary, born as the rest of men are; that he was but a mere man, but that, on account of his great virtue, the
Almighty adopted him as his Son (9).
\section{Saturninus and Basilides}
Saturninus and Basilides were disciples of Menander, whose history we have already seen; and they
made some additions to the heresy of their master. Saturninus, a native of Antioch, taught, with
Menander, as Fleury tells us (10), that there was one only Father, unknown to all, who created the angels,
and that seven angels created the world and man. The God of the Jews, he said, was one of these
rebellious angels, and it was to destroy him that Christ appeared in the form of man, though he never
had a real body. He condemned matrimony and procreation as an invention of the devil. He attributed
the Prophecies partly to the angels, partly to the devil, and partly to the God of the Jews. He also said,
according to St. Augustine (Heres. iii), that the Supreme Virtue that is, the Sovereign Father having
created the angels, seven of them rebelled against him, created man, and for this reason : Seeing a celestial
light, they wished to retain it, but it vanished from them; and they then created man to resemble it,
saying, " Let us make man to the image and likeness." Man being thus created, was like a mere worm,
incapable of doing anything, till the Sovereign Virtue, pitying his image, placed in him a spark of himself,
and gave him life. This is the spark which, at the dissolution of the body, flies to heaven. Those of his sect
alone, he said, had this spark; all the others were deprived of it, and, consequently, were reprobate.\\
(9) N. Alex. loc. cit, art. 6; Fleury, loc. cit. n, 42. [N.B Fleury puts Ebion first, next Cerinthus, and lastly
Menander,]\\
(10) Fleury, n. 19.

Basilides, according to Fleury, was a native of Alexandria, and even exceeded Saturninus in fanaticism.
He said that the Father, whom he called Abrasax, produced Nous, that is, Intelligence; who produced
Logos, or the Word; the Word produced Phronesis, that is, Prudence; and Prudence, Sophia and
Dunamis, that is, Wisdom and Power. These created the angels, who formed the first heaven and other
angels; and these, in their turn, produced a second heaven, and so on, till there were three hundred and
sixty-five heavens produced, according to the number of days in the year. The God of the Jews, he said,
was the head of the second order of angels, and because he wished to rule all nations, the other princes
rose up against him, and, on that account, God sent his first-born, Nous, to free mankind from the
dominion of the angels who created the world. This Nous, who, according to him, was Jesus Christ, was
an incorporeal virtue, who put on whatever form pleased him. Hence, when the Jews wished to crucify
him, he took the form of Simon the Cyrenean, and gave his form to Simon, so that it was Simon, and not
Jesus, who was crucified. Jesus, at the same time, was laughing at the folly of the Jews, and afterwards
ascended invisibly to heaven. On that account, he said, we should not venerate the crucifix, otherwise we
would incur the danger of being subject to the angels who created the world. He broached many other
errors; but these are sufficient to show his fanaticism and impiety. Both Saturninus and Basilides fled
from martyrdom, and always cloaked their faith with this maxim " Know others, but let no one know
you." Cardinal Orsi says (11) they practised magic, and were addicted to every species of incontinence,
but that they were careful in avoiding observation. They promulgated their doctrines before Menander,
in the year 125; but, because they were disciples of his, we have mentioned them after him.\\
(11) Orsi, t. 2, I. 3, n. 23.
\section{The Nicholites}
The Nicholites admitted promiscuous intercourse with married and single, and, also, the use of meats
offered to idols. They also said that the Father of Jesus Christ was not the creator of the world. Among the
other foolish doctrines they held, was one, that darkness, uniting with the Holy Ghost, produced a matrix
or womb, which brought forth four Eons; that from these four Eons sprung the evil Eon, who created the
Gods, the angels, men, and seven demoniacal spirits, This heresy was of short duration; but some new
Nicholites sprung up afterwards in the Milanese territory, who were condemned by Pope Nicholas II.
The Nicholites called themselves disciples of Nicholas the Deacon, who, according to Noel Alexander,
was esteemed a heresiarch by St. Eusebius, St. Hilarian, and St. Jerome. However, Clement of Alexandria,
Eusebius, Theodoret, Baronius, St. Ignatius the Martyr, Orsi, St. Augustine, Fleury, and Berti, acquit him
of this charge (12).\\
(12) Nat. Alex. t. 5, diss. 9; Baron. An. 68, n. 9; Orsi, t. 1, n. 64; Fleury, t. 1, L 2, n. 21; Berti, loc. cit.

\chapter{Heresies of the Second Century}
\section{Copocrates}
Corpocrates was a native of Alexandria, or, as others say, of Samosata. His followers were called
Gnostics that is, learned or enlightened. He said that Jesus Christ was the son of Joseph, born as other
men are, and distinguished from them by his virtue alone, and that the world was created by angels.
Another blasphemous doctrine of his was, that, to unite ourselves with God, we should practise all the
unclean works of concupiscence; our evil propensities should be followed in everything, for this, he said,
was the enemy spoken of in the Gospel (1), to which we should yield, and, by this means, we show our
contempt for the laws of the wicked angels, and acquire the summit of perfection; and the soul, he said,
would pass from one body to another till it had committed all sorts of unclean actions. Another of his
doctrines was, that every one had two souls, for without the second, he said, the first would be subject to
the rebellious angels. The followers of this hellish monster called themselves Christians, and, as a
distinctive mark, they branded the lower part of the ear with a red iron. They paid the same veneration to
the images of Pythagoras, Plato, and the other philosophers, as to that of Jesus Christ. Corpocrates lived
in the year 160.
\section{Valentine}
Valentine, who, it was supposed, was an Egyptian, separated himself from the Church, because he was
disappointed in obtaining a bishopric. He came to Rome in 141, and abjured his errors, but soon again
embraced them, and persevered in them till his death (2). He invented a fabulous genealogy of Eons or
Gods; and another of his errors was, that Jesus Christ did not become incarnate in the womb of the Virgin
Mary, but brought his body from heaven. He admitted in man a continual exercise of spirit, which,
uniting with the flesh, rendered lawful every sensual pleasure; and he divided mankind into three classes
the carnal, the animal, and the spiritual. His followers, he said, were the spiritualists, and, on that
account, were exempt from the necessity of good works, because, having arrived at the apex of
perfection, and being certain of eternal felicity, it was useless for them to suffer, or observe the law. The
carnal, he said, were excluded from eternal salvation and predestined to hell (3).\\
(1) N. Alex. t. 6, c. 3, ar. 2; Flemy, l. 3, n. 20; Berti, t. 1, c. 3; Bernin. t. 1, c. 2. \\
(2) Van Ranst, His. p. 20.\\
(3) Fleury, t. 1, l. 3, n. 2627; Bernin. t. 1, c. 5; Graveson, t. 3, . 49; N. Alex. t. 6, c. 3, ar. 6.\\

Three sects take their origin from Valentine. The first were called Sethites : These paid such honour to
Seth, that they said Jesus Christ was born of him, and some went so far as to say that Jesus Christ and
Seth were one and the same person. The second sect were called Cainites : These venerated as saints all
those who the Scripture tells us were damned as Cain, Core, the inhabitants of Sodom, and especially
Judas Iscariot. The third were called Ophites : These said that Wisdom became a serpent, and; on that
account, they adored Jesus Christ as a serpent; they trained one of these reptiles to come out of a cave
when called, and creep up on the table where the bread for sacrifice was placed; they kissed him while he
crept round the bread, and, considering it then sanctified by the reptile, whom they blasphemously called
Christ, they broke it to the people, who received it as the Eucharist (4).

Ptolemy and Saturninus were disciples of Valentine; but their master admitted thirty Eons, and they
added eight more. He also had other disciples : Heraclion, whose followers invoked over the dead certain
names of principalities, and anointed them with oil and water; Marcus and Colarbasus taught that all
truth was shut up in the Greek alphabet, and, on that account, they called Christ Alpha and Omega (5);
and Van Ranst adds to the list the Arconticites, who rejected the sacraments Florinus, who said that God
was the author of sin and Blastus (6), who insisted that Easter should be celebrated after the Jewish
fashion. The disciples of Valentine made a new Gospel, and added various books to the Canon of the
Scriptures, as \textbf{" The Parables of the Lord," " The Prophetic Sayings and the Sermons of the Apostles."} It is
needless to add that all these were according to their own doctrines.
\section{Epiphanes}
Epiphanes, the son of Carpocrates, besides defending the damnable opinions of his father, openly
rejected the law of Moses, and especially the two last precepts of the Decalogue. He also rejected the
Gospel, though he pretended to follow it (7).\\
(4) Fleury, t. 1, 1. 3, n. 30; Bernin. t. 1, c. 2; Van Ranst, p. 20.\\
(5) Fleury, l. 3, n. 30, l. 4, n. 9 \& 10. \\
(6) Van Ranst, p. 22. (7) Fleury, l. 3, n. 20; Bern. t. 1, c. 2.
\section{Prodicus}
Prodicus taught that it was lawful to deny the faith to avoid death; he rejected the worship of an
invisible God, and adored the four elements and the sun and the moon; he condemned all prayers to God
as superstitious, but he prayed to the elements and the planets to be propitious to mankind (8). This
impious worship he always performed naked. Noel Alexander and Theodoret assign to this heretic the
institution of the sect called Adamites; these always performed their religious exercises in their churches,
or rather brothels, as St. Epiphanius calls them, naked, pretending by this to imitate the innocence of
Adam, but, in reality, practising every abomination (9).
\section{Tatian}
Tatian was born in Assyria, and was a disciple of St. Justin Martyr. He was the founder of the sect
called Encratics, or Continent; he taught, with Valentine, that matter was uncreated and eternal; he
attributed the creation to God, but through the instrumentality of an inferior Eon, who said let there be
light, not by way of command, but of supplication, and thus light was created. He denied, with Valentine,
the resurrection of the dead, and human flesh, he said was too unworthy to be united with the divinity in
the person of Christ. He deprived man of free will, saying he was good and spiritual, or bad and carnal,
by necessity, according as the seed of divine grace was infused or not into him; and he rejected the law of
Moses, as not instituted by God, but by the Eon who created the world. Finally, he condemned
matrimony, prohibited the use of flesh-meat and wine, and, because he used nothing but water in the
consecration of the chalice, his disciples were called Hydroparastati, or Aquarii (10).
\section{Severus}
Severus was a disciple of Tatian; but differed from his master in some essential points, especially in
admitting the law of Moses, the Prophets, and the Gospels. Julius Capianus, a disciple of Valentine,
joined with Severus, and was the founder of the heresy of the Doceti, who said that Jesus had not a real,
but an apparent, body. He wrote a book on continence, in which he quoted a passage of the spurious
gospel used by the Egyptians, in which Jesus Christ is made to curse matrimony. In his commentaries on
Genesis he says marriage was the forbidden fruit (11).\\
(8) Bern. loc. cit. \\
(9) N. Alex. t. 6, c. 3, ar. 12; Gotti, Ver. Eel. t. 2, c, 27, s. 1; Bernin. loc. cit.\\
(10) Orsi, t. 2, l. 4, n. 11; Fleury, t, 1, l. 4, n. 8; Baron. An. 174, n.3, 4; N. Alex. t. 6, c. 3, ar. 7. \\
(11) Fleury, loc. cit. n. 8; Orsi, loc, cit. n. 12.
\section{Cerdonius}
Cerdonius followed the doctrines of Simon, Menander, and Saturninus; besides, he taught, with
Maims, the existence of two first principles, or Gods, a good and a bad one, and admitted the resurrection
of the soul, but not of the body. He rejected all the Gospels, except St. Luke’s, and mutilated that in
several places (12)
\section{Marcion}
Marcion was a native of the city of Sinope, in the province of Pontus, and the son of a Catholic bishop.
In his early days he led a life of continence and retirement; but for an act of immorality he was cut off
from the Church by his own father. He then went to Rome, and endeavoured to accomplish his
restoration; but not being able to succeed, he, in a fit of rage, said " I will cause an eternal division in
your Church." He then united himself to Cerdonius, admitting two principles, and founding his doctrine
on the sixth chapter of St. Luke, where it is said a good tree cannot bring forth bad fruits. The good
principle, he said, was the author of good, and the bad one of evil; and the good principle was the father
of Jesus Christ, the giver of grace, and the bad one, the creator of matter and the founder of the law. He
denied the incarnation of the Son of God, saying it was repugnant to a good God to unite himself with the
filthiness of flesh, and that his soul should have for a companion a body infected and corrupt by nature.
He also taught the existence of two Gods one, the good God; the other, an evil one, the God of the Jews,
and the creator of the world. Each of these Gods promised to send a Christ. Our Christ appeared in the
reign of Tiberius, and was the good Christ; the Jewish Christ did not yet come. The Old Testament he
rejected, because it was given by the bad principle, or God of the Jews. Among other errors, he said, that
when Jesus descended into hell, he did not save Abel, or Henoc, or Noah, or any other of the just of the
old law, because they were friends of the God of the Jews; but that he saved Cain, the Sodomites, and the
Egyptians, because they were the enemies of this God (13).
(12) Fleury, l. 3, n. 30; Nat. Alex. t. 6, c. 3, ar. 4; Orsi, t. 2, l. 3, n. 44. 
(13) Orsi, t. 2, L 3, n. 45; N. Alex. t. 6, c. 3, ar. 5; Baron. Ann. 146, n. 9, \&c.; Fleury, t. 1, l. 3, n. 34.
\section{Apelles}
Apelles, the most famous disciple of Marcion, was excommunicated by his master for committing a
crime against chastity, and felt his disgrace so much that he fled to Alexandria. This heretic, among other
errors, said that God created a number of angels and powers, and among the rest a power called the
Lord, who created this world to resemble the world above, but not being able to bring it to perfection, he
repented him of having created it (14). Van Ranst says that he rejected the Prophecies, and said the Son of
God took a body of air which, at his ascension, dissolved into air again.
\section{Montanus}
Montanus, as Cardinal Orsi tells us (15), was born in Ardraba, an obscure village of Mysia. He first led
such a mortified life that he was esteemed a saint; but, possessed by the demon of ambition, his head was
turned. He began to speak in an extraordinary manner, make use of unknown words, and utter
prophecies in contradiction to the traditions of the Church. Some thought him possessed by a spirit of
error; others looked on him as a saint and prophet. He soon acquired a number of followers, and carried
his madness to the utmost excess; among others who joined him were two loose women of the names of
Prisca or Priscilla and Maximilla, and, seemingly possessed by the same spirit as himself, they uttered the
most extraordinary rhodomontades. Montanus said that he and his prophetesses received the plenitude
of the Holy Ghost, which was only partially communicated to others, and he quoted in his favour that
text of St. Paul (I. Corinthians, xiii, 9), " By part we know, and by part we prophesy ;" and they had the
madness to esteem themselves greater than the apostles, since they had received the Holy Ghost
promised by Jesus Christ in perfection. They also said that God wished, at first, to save the world, by
means of Moses and the prophets; when he saw that these were not able to accomplish it, he himself
became incarnate; but even this not sufficing, he descended in the Holy Ghost into Montanus and his
prophetesses. He established nine fasting-days and three Lents in the year. Among other errors he
prohibited his disciples to fly from persecution, and refused to admit sinners to repentance, and
prohibited second marriages (16).\\
(14) Fleury, loc. cit. . 35, \\
(15) Orsi, t. 2, l. 4, n. 17. \\
(16) Euseb. Hist. Eccl. l. 5, c. 15.\\

Eusebius tells us that he died miserably, having hanged himself (17)
\section{Cataphrigians, Artotirites, Peputians, Ascodrogites, Pattalorinchites}
The heresy of Montanus shot forth different branches, as the Cataphrigians, Artotirites, Peputians,
Ascodrogites, and Pattalorinchites. The Cataphrigians were called from the nation to which Montanus
belonged. The Eucharistic bread they used was made of flour and blood taken from the body of an infant
by puncturing it all over; if the infant died he was considered a martyr, but if he survived he was
regarded as high priest. This we learn from Noel Alexander (18). The Artotirites were so called, because
in the sacrifice of the Eucharist, they offered up bread and cheese. The Peputians took their name from an
obscure village of Phrigia, where they held their solemn meetings; they ordained women priests and
bishops, saying there was no difference between them and men. The Ascodrogites were no better than the
ancient bacchanalians; they used bottles which they filled with wine near the altars, saying that these
were the new bottles Jesus Christ spoke of " They shall put new wine into new bottles, and both are
preserved." The Pattalorinchites were so called, because they wore a small stick in the mouth or nose, a
sign of strict silence; they were so called, from pattalos, a stick, and rinchos, the nose (19).
\section{Bardesanes}
Bardesanes, a native of Edessa, in Syria, lived in this age also. He was celebrated in the time of Marcus
Aurelius for his learning and constancy in defending the faith. He told the Philosopher Apollonius, the
favourite of the Emperor, who endeavoured to pervert him, that he was ready to seal his belief with his
blood. He opposed the errors of Valentine; but, being educated in his school, he was infected with some
of them, especially disbelieving the resurrection of the dead. He wrote many works in refutation of the
heresies of his day, especially an excellent treatise on fate, which St. Jerome, in his catalogue of
ecclesiastical writers, praises highly. We may truly say, with Noel Alexander, that the fall of so great a
man is to be lamented (20).\\
(17) Baron. An. 173, n. 20; N. Alex. t. 6, sec. 2, c. 3, ar. 8; Fleury, t. 1, 1. 4, n. 5; Bernin. t. 1, c. 8; Orsi, t. 2, L
4, n. 18.\\
(18) Nat. Alex. cit. ar. 8, n. 11; St. Angus. \& St. Cyril. [St. Epiphanius says it is the Peputians.]\\
(19) Van Ranst, His. Heres. p. 24; Vedia anche Nat. Alex. loc. cit.\\
(20) Nat. Alex. t. 6, c. 3, ar. 9; Van Ranst, p. 24.\\
\section{Theodotus the Currier, Artemon, and Theodotus Argentarius}
Theodotus the Currier, so called on account of his trade, was a native of Byzantium, and he, along
with Artemon, asserted like Ebion and Cerinthus, that Christ was mere man. Besides this there was 
another Theodotus, called Argentarius, or the Banker, who taught that Melchisadech was Christ, or even
greater than Christ, on account of that verse of the Psalms " Thou art a priest for ever, according to the
order of Melchisadech ;" and his followers were afterwards called Melchisadechites (21).
\section{Hermogenes}
Hermogenes said that matter was uncreated and eternal. Tertullian, Eusebius, and Lactanctius refuted
this error. He also taught that the devils would hereafter be united with matter and that the body of Jesus
Christ was in the sun (22).\\
(21) N. Alex. loc. cit. ar. 10; Fleury, f. 1, l. 4, n. 33, 34. \\
(22) Fleury, loc. cit. n. 21; N. Alex, loc. cit. ar. 15.\\

\chapter{Heresies of the Third Century}
\section{Praxeas}
 Praxeas, a native of Phrigia, was at first a Montanist, but afterwards becoming an enemy of Montanus,
he caused him to be condemned by Pope Zepherinus, concealing his own heresy at the same time. Being
soon discovered, he retracted his opinions, but soon afterwards openly proclaimed them. He denied the
mystery of the Trinity, saying that in God there was but one person and one nature, which he called the
Father. This sole person, he said, descended into the womb of the Virgin, and being born of her by means
of the incarnation, was called Jesus Christ. According to this impious doctrine, then, it was the Father
who suffered death, and on that account his followers were called Patripassionists. The most remarkable
among his disciples were Berillus, Noetus, and Sabellius. Berillus was Bishop of Bostris in Arabia; he said
that Christ, before his incarnation, had no divinity, and in his incarnation had no divinity of his own, but
only that of the Father. Noel Alexander says that Origen refuted him, and brought him back to the
Catholic faith (1). Noetus, more obstinate in error, said that the Father, the Son, and the Holy Ghost were
but one person and one God; he and his followers were cut off from the Church, and, as he died
impenitent, he was refused Christian burial (2). The most celebrated promoter of this error was Sabellius.
\section{Sabellius}
Sabellius was born in the Ptolemais in Africa, and lived in the year 227. He shed a greater lustre, if we
may say so, on the heresy of his master, and on that account this impious sect was called Sabellians. He
denied the distinction of the three persons in the Trinity, and said they were but three names to
distinguish the different operations of the Divinity. The Trinity, he said, was like the sun, in which we
distinguish the light, the heat, and the form, though the sun be but one and the same. The light represents
the Son, the heat the Holy Ghost, and the figure or substance of the sun itself the Father, who, in one
person alone, contained the Son and the Holy Ghost (3). This error we will refute in the last part of the
work.\\
(1) Nat. Alex. t. 7, s. 3, c. 3, ar. 1, ex Euseb.; Van Ranst, p. 65.\\
(2) Nat, Alex, ibid, c. 3, ar. 7; Van Ranst, p. 48.\\
(3) Nat. Alex. t. 7, c. 3, ar. 7; Orsi, t. 2, l. 5, n. 14; Hermant, 1. 1, c. 60; Fleury, I. 7, n. 35.
\section{Paul of Samosata}
Paul of Samosata was Bishop of Antioch. Before his appointment to the see he was poor, but
afterwards, by extortion and sacrilege, by selling justice, and making false promises, he amassed a great
deal of wealth. He was so vain and proud that he never appeared in public without a crowd of courtiers;
he was always preceded by one hundred servants, and followed by a like number, and his own praises
were the only subjects of his sermons; he not only abused those who did not flatter him, but frequently
also offered them personal violence; and at length his vanity arrived at such a pitch that he had a choir of
courtezans to sing hymns in his praise in the church; he was so dissolute in his morals that he had always
a number of ladies of lax morals in his train. In fine, this impious prelate crowned all his crimes with
heresy. The first of his blasphemies was, that Jesus Christ never existed until he was born of the Virgin,
and hence he said he was a mere man; he also said that in Jesus there were two persons and two sons of
God, one by nature and the other by adoption; he also denied the Trinity of the Divine persons, and
although he admitted the names of the Father, and of the Son, and of the Holy Ghost, not, however,
denying, as Orsi thinks, personal existence to the Son and the Holy Ghost, yet he did not recognize either
one or the other as persons of the Trinity, attributing to the Father alone the incarnation and passion (4).
His disciples inserted those errors in their profession of faith, and in the formula of Baptism, but N.
Alexander says that it is uncertain whether Paul was the author of this heresy.
\section{Manes}
Manes was the founder of the Manicheans, and he adopted this name on account of taking to himself
the title of the Paraclete, and to conceal the lowliness of his condition, since he was at first only a slave in
Persia, but was liberated and adopted by an old lady of that country. She sent him to the public academy
to be educated, but he made little progress in learning. Whatever he wanted in learning he made up in
impudence, and on that account he endeavoured to institute a new sect; and, to enlist the peasantry
under the banner of his heresy, he studied magic with particular attention. To acquire a name for himself
he undertook to cure the King of Persia’s son, who was despaired of by the physicians. Unfortunately for
him, however, the child died, notwithstanding all his endeavours to save him, and he was thrown into
prison, and would have been put to death only he bribed the guards to let him escape. Misfortune,
however, pursued him : after travelling through various countries, he fell again into the King’s hands,
who ordered him to be flayed alive with a sharp-pointed reed; his body was thrown to the beasts, and his
skin hung up in the city gate, and thus the impious Manes closed his career. He left many followers after
him, among whom was St. Augustine, in his youth, but, enlightened by the Almighty, he abandoned his
errors, and became one of his most strenuous opponents (5).\\
(4) Orsi, t. 3, l. 8, n. 15; Gotti de Vera Rel. t. 2, c, 11, s. 2; N. Alex. t. 7, c. 3, ar. 8, sec. 2; Hermant, t. 1, c. 63;
Fleury, t. 2, l. 8, n. 1.\\
(5) Baron. Ann. 277, ex n. 1; Nat. Alex. t. 7, c. 3, ar. 9, sec. 1.\\

The errors of Manes can be classed under the following heads : 1st. He admitted the plurality of Gods,
alleging that there were two principles, one of good and the other of evil. Another of his errors was, that
man had two souls one bad, which the evil principle created, together with the body, and another, good,
created by the good principle, which was co-eternal, and of the same nature with God. All the good
actions which man performs he attributes to the good soul, and all the evil ones he commits to the bad
soul. He deprived man of free-will, saying that he was always carried irresistibly forward by a force
which his will could not resist. He denied the necessity of baptism, and entirely abolished that sacrament.
Among many other errors, the Manicheans detested the flesh, as being created by the evil principle, and,
therefore, denied that Jesus Christ ever took a body like ours, and they were addicted to every sort of
impurity (6). They spread almost over the entire world, and though condemned by many Popes, and
persecuted by many Emperors, as Dioclesian, Gratian, and Theodosius, but especially by Justin and
Justinian, who caused many of them to be burned alive in Armenia, still they were not annihilated till the
year 1052, when, as Baronius relates, Henry II., finding some of them lurking in France, caused them to be
hanged. The refutation of this heresy we have written in the book called the Truth of the Faith (7).
\section{Tertullian}
Tertullian was born, as Fleury (8) relates, in Carthage, and his father was a centurion in the Pretorian
Bands. He was at first a Pagan, but was converted about the year 197, and was a priest for forty years,
and died at a very advanced age. He wrote many works of the highest utility to the Church, on Baptism,
Penance, Idolatry, on the Soul, on Proscriptions, and an Apology for the Christians, which has acquired
great celebrity. Although in his book on Proscriptions he calls Montanus a heretic, still, according to the
general opinion of authors, he fell into Montanism himself. Baronius says that he was cut off from the
Church, and excommunicated by Pope Zepherinus (9). Tertullian was a man of the greatest austerity; he
had the greatest veneration for continence; he practised extraordinary watchings, and on account of a
dispute he had with the clergy of Rome, he attached himself to the Montanists, who, to the most rigid
mortification, joined the belief that Montanus was the Holy Ghost. N. Alexander proves, on the authority
of St. Jerome, St. Hilary, St. Pacianus, St. Optatus, and St. Augustine, that he asserted the Church could
not absolve adulterers, that those who married a second time were adulterers, and that it was not lawful
to fly from persecution. He called the Catholics, Psichici, or Animals. Fleury says (10), that Tertullian
taught that the soul was a body, of a palpable form, but transparent, because one of the Prophetesses
heard so in a vision. Both Fleury and Noel Alexander say (11), that he forsook the Montanists before his
death, but a sect, who called themselves Tertullianists after him, remained in Carthage for two hundred
years, until the time of St. Augustine, when they once more returned to the bosom of the Church.
(6) Nat. Alex ibid, vide sec. 2; Hermant, t. 1, c. 65; Fleury, t. 2, L 8, n. 1012; Baron. Ann. 277, . 1, \& seq.;
Graves, in sec. 3.\\
(7) Verità della Fede, part 3, c. 2, sec. 2.\\
(8) Fleury, t. 1, L 4, n. 47.\\
(9) Baron. Ann. 201, n. 3, \& seq. ad. 11; Fleury, t. 1, l. 25 \& 26; Orsi,t. 3, l. 8, n. 28.\\
(10) Fleury, t. I, 1. 5, n. 25\\
(11) Fleury, t. 1, I. 6, n. 3, cum St. Augus. \& Nat. Alex. t. 6, c. 3, ar. 8, n. 9.
\section{Origen}
Origen was an Egyptian, and his early days were spent in Alexandria. His father was St. Leonidas the
Martyr, who had him educated in every branch of sacred and profane literature (12). It is said his own
father held him in the highest veneration, and that often while he slept he used to kiss his bosom, as the
temple where the Holy Ghost dwelt (13). At the age of eighteen he was made Catechist of the Church of
Alexandria, and he discharged his duties so well that the very pagans flocked to hear him. Plutarch, who
afterwards became an illustrious martyr of the faith of Christ, was one of his disciples. In the height of the
persecution he never ceased to assist the confessors of Christ, despising both torments and death. He had
the greatest horror of sensual pleasures, and it is related of him that for fear of offending against chastity,
and to avoid temptation, he mutilated himself, interpreting the 12th verse of the 19th chapter of St.
Matthew in a wrong sense (14). He refuted the Arabians, who denied the immortality of the soul, and
converted Berrillus, as we have already seen, who denied the divinity of Jesus Christ. He also converted
Ambrose from the errors of the Valentinians. He was so desirous of martyrdom, that his mother was
obliged to take away his clothes, to prevent him from going to his father, who was in prison for the faith.
All this, however, was to no purpose; he avoided her vigilance, flew to his father, and when he would not
be allowed to speak to him, he exhorted him by letter to persevere in the faith. At the age of eighteen he
was Prefect of the studies of Alexandria. When he was composing his Commentaries on the Scriptures, he
dictated to seven or eight amanuenses at the same time. He edited different editions of the Scriptures,
compiling the Tetrapla, the Hexapla, and the Octapla, The Tetrapla had four columns in each page; in the
first was the version of the seventy, or Septuagint, in the second that of Aquila, in the third that of
Simmachus, and in the fourth that of Theodotian. The Hexapla had six columns, and, besides the former,
contained the Hebrew text and a Greek translation. Finally, the Octapla contained, besides the former,
two other versions, compiled by some Hebrews. His name was so famous at that time that all the priests
and doctors consulted him in any difficult matter. Presuming too much on his wisdom, he fell into
different errors, by wishing to interpret many texts of Scripture in a mystical, rejecting the literal, sense.
Those, he says, who adhere to the letter of the Scripture will never see the kingdom of God (15), hence we
should seek the spirit of the word, which is hidden and mysterious. He is defended by some; but the
majority condemn him, although he endeavoured to clear himself by saying that he wrote his sentiments
merely as opinions, and subjected them to the judgment of his readers (16).\\
(12) Nat. Alex. t. 1, ar. 12.\\
(13) Fleury, I. 5, n. 2; Orsi, l. 5, n. 27.\\
(14) Nat. Alex. t. 7, nr. 12.\\
(15) Origen, Stromata, l. 10.\\
(16) Orsi, l. 6, n. 61.\\

He was obliged to go into Achaia, a country at that time distracted by various heresies. In his journey he
persuaded two bishops of Palestine whom he visited, that it would be of great service to the Church if he
was ordained priest (17). Yielding to his suggestions they ordained him, and this so displeased
Demetrius, Bishop of Alexandria, that in a council he deposed and excommunicated him. Several other
bishops, however, received him in his misfortunes, and entertained him honourably. Orsi, on the
authority of Eusebius, tells us (18), that, in the persecution of Decius he was imprisoned a long time,
loaded with irons, and a great iron ring on his neck; and that he was not only tortured in the legs in a
horrible manner, but was likewise put on the rack. Dionisius, Eusebius says (19), wrote him a letter, or
rather a small treatise,, to animate and console him; and from that circumstance, Cardinal Orsi (20)
proves the fallacy of Du Pin’s conjecture, that the sentence passed against him by Demetrius, was
enforced under his successors Aracla and Dionisius. Origen did not long survive the torments he endured
in that persecution. He died in Tyre, in the year 253, the sixty-ninth of his age (21).

Bernini tells us, on the authority of St. Epiphanius (22), (thinking, however, that this was foisted into St.
Epiphanius’s works by the enemies of Origen) that he denied the faith by offering incense to idols, to
avoid the indignities and insults inflicted on him by an Ethiopian, and that he was then freed from
prison, and his life spared. After that he went from Alexandria to Jerusalem, and at the request of the
clergy and people went into the pulpit to preach. It happened, however, that opening the book of the
Psalms, to explain them, the first words he read were those of the 49th Psalm : " God said to the sinner,
why dost thou declare my justices and take my covenant into thy mouth ? " Struck dumb with sorrow, he
began to weep bitterly, and left the pulpit without saying a word. Not only St. Epiphanius, but Eusebius
(23) before him, bear witness to Origen’s fall. Although Bernini (24) says this story is quite fabulous, yet 
Petavius, Daniel Uerius, Pagi, and especially Noel Alexander (25), say it is a fact. Roncaglia (26) is of
opinion that Noel Alexander’s arguments are groundless, and that Baronius’s opinion carries more
weight with it.\\
(17) Nat. Alex, ibid; Orsi, n. 30.\\
(18) Orsi, t. 3, l. 7, n. 33. \\
(19) Euseb. His. Eccl. l. 6. \\
(20) Orsi. t. 3, I. 7, n. 33.\\
(21) Orsi, loc. cit.; Hermant, t. 1, c. 68; Bar. Ann. 204, n. 8; V. Ranst, p. 42; Graves, s. 3. \\
(22) Bernin. Istor. t.1, c. 1, p. 125. \\
(23) Euseb. l. 6; Hist. Eccl. c. 59.\\
(24) Baron. Ann. 253, n. 117, \& seq. cum Graves, loc. cit. \\
(25) Petav. in Animadv. in St. Epiph. Heres. 64; Huetius, l. 1; Orig. c. 4; Pagius ad an. 251, n. 19; Nat. Alex. t. 7, diss. 15, q. 2, art, unic. \\
(26) Rone. not. in Natal, loc. cit. \\

We can decide nothing as to the salvation of Origen, though Baronius says that St. Simeon Salus saw him
in hell; still, all is a mystery known to God alone. We know, however, on the authority of Baronius, that
his doctrine was condemned by Pope Anastasius and Pope Gelasius, and afterwards by the fifth general
council (27).

The substance of the errors of Origen, as well as I could collect from the works of Noel Alexander, Fleury,
Hermant, Orsi, Van Ranst (who gives a great deal of information in a small space), and others, was all
included in his Periarchon, or Treatise on Principles. This treatise, Fleury says, was translated by Rufinus,
who endeavoured to correct it as much as possible. The intent of Origen in this work was to refute
Valentine, Marcion, and Ebion, who taught that men are either essentially good or essentially wicked. He
said that God alone was good and immutable, but that his creatures were capable of either good or evil,
by making use of their free will or a good purpose, or perverting it for a wicked one. Another of his
opinions was that the souls of men were of the same nature as the celestial spirits, that is, composed of
spirit and matter; that they were all created before the beginning of the world, but that, as a punishment
for some crimes committed, they were shut up in the sun, moon, and other planets, and even in human
bodies, as it were in a prison, to punish them for a time; after which, being freed from their slavery by
death, they went to heaven to receive the reward of their virtues, or to hell to suffer the punishment of
their sins, but such rewards and punishments were not eternal. Hence, he said, the blessed in heaven
could be banished from that abode of happiness for faults committed there, and that the punishment of
the devils and the damned would not last for all eternity, because at the end of the world Jesus Christ
would be again crucified, and they would participate in the general redemption. He also said that before
the creation of this world there existed many others, and that after this had ceased to exist many more
would be created, for, as God was never idle, so he never was without a world.\\
(27) Baron. Ann. 400, \&c.

He taught many other erroneous opinions; in fact his doctrine is entirely infected with the maxims of
Plato, Pythagoras, and the Manicheans. Cassiodorus, speaking of Origen, says, I wonder how the same
man could contradict himself so much; for since the days of the Apostles he had no equal in that part of
his doctrine which was approved of, and no one ever erred more grossly in the part which was
condemned. Cabassutius (28) says, that Pope Gelasius, following the example of Anastatius, gave this
sentence relative to Origen in the Roman council : "We declare that those works of Origen which the
blessed Jerome does not reject can be read, but we condemn all others with their author."

After the death of Origen his followers disturbed the Church very much by maintaining and propagating
his errors. Hermant (29) relates that Pope Anastasius had a great deal of difficulty in putting down the
troubles occasioned by the Origenists in Rome, who got footing there under the auspices of Melania, by
means of the priest Rufinus. The author of the notes on Floury, says, that Anastasius wrote to John of
Jerusalem to inform him of how matters were going on, and that he, on that account, cut off Rufinus from
the Church. In the reign of the Emperor Justinian, some Origenist monks who lived in a laura founded by
St. Saba, under the abbot Nonnus, began to disseminate their errors among this brethren, and in a short
time infected the principal laura, but were expelled by the abbot Gelasius. Favoured, however, by
Theodore of Cesarea, they got possession of the great laura again, and expelled the greater part of the
monks who disagreed with them. In the meantime, Nonnus died, and his successor George being
deposed for immorality by his own party, the Catholic monks again got possession of the laura, and
elected Conon, one of this party, abbot (30). Finally, in the twelfth canon of the second council of
Constantinople, both Origen and all those who would persist in defending his doctrine were condemned
(31).\\
(28) Cabassut. Notit. Hist. Cone. Constan. II. an. 553, n. 14. in fin.. \\
(29) Hermant, t. 1, c, 132. \\
(30) Orsi, t. 18, l. 41, n. 1 \& 5, ad 7,\\
(31) Orsi, al luogo cit. n. 70
\section{Novatus and Novatian}
Novatus and Novatian. Novatus was a priest of the Church of Carthage. St. Cyprian relates that he was
a man of a turbulent disposition, seditious and avaricious, and that his faith was suspected by the
bishops. He was accused of robbing the orphans and widows, and appropriating to his own use the
money given him for the use of the Church. It is said he allowed his father to die of starvation, and
afterwards refused to bury him; and that he caused the death of his wife by giving her a kick, and causing
premature labour. He was also one of the principal agents in getting the deacon Felicissimus ordained
priest without the leave or knowledge of St. Cyprian, his bishop, and was one of the principal leaders of
the schism of Novatian, exciting as many as he could to oppose the lawful Pope, Cornelius (32).

We now come to speak of the character and errors of Novatian. Being possessed by an evil spirit he was
baptized in bed during a dangerous fit of sickness, and when he recovered he neglected getting the
ceremonies of baptism supplied, and never received confirmation, which, according to the discipline of
the Church in those days, he ought to have received after baptism, and his followers, for that reason,
afterwards rejected this sacrament. He was afterwards ordained priest, the bishop dispensing in the
irregularity he incurred by being baptized in bed. Hence his ordination gave great umbrage both to the
clergy and people. While the persecution was raging the deacons begged of him to leave his place of
concealment, and assist the faithful, who were dragged to the place of punishment; but he answered, that
he did not henceforward intend to discharge the duties of a priest; that he had his mind made up for
other objects. This was nothing less than the Popedom, which he had the ambition to pretend to, puffed
up by the applause he received for his oratorical powers. At this time, Cornelius was elected Pope, and
he, by intrigue, got himself consecrated privately by three ignorant bishops whom he made intoxicated.
Thus he was the first anti-Pope who ever raised a schism in the Church of Rome. But what will not
ambition do ? While he administered the Eucharist to his partizans, he exacted an oath from each of them,
saying, " Swear to me, by the blood of Jesus Christ, that you will never leave my party and join Cornelius"
(33). \\
(32) Baron. An. 254, n. 50.; Nat. t. 7, c. 3, or. 3, 4; Fleury, t. 1, 1. 6, n. 51. \\
(33) Nat, loc. cit.; Baron, n. 61, \&c. \\

The errors of Novatus and Novatian were the following: they denied that the Church could use any
indulgence with those who became idolaters through fear of persecution, or that she could grant pardon
for any mortal sin committed after baptism, and they denied the sacrament of confirmation. Like the
Montanists, they condemned second marriages, and refused communion on the point of death to those
who contracted them (34).
\section{Nipos, The Angelicals and the Apostolicals}
These were not the only heretics who disturbed the Church during this century. Nipos, an Egyptian
bishop, about the year 284, again raked up the errors of the Millenarians, taking the promise of the
Apocalypse in a literal sense, that Jesus Christ would reign on earth for the space of a thousand years,
and that the saints should enjoy all manner of sensual delights. The Angelicals offered the supreme
adoration which should be given to God alone, to the angels; adored them as the creators of the world,
and pretended to lead angelic lives themselves.

The Apostolicals said it was not lawful for any one to possess property of any sort, and that the riches of
this life were an insurmountable obstacle to salvation. These heretics received no married persons into
this sect (35).\\
(34) Nat. Alex, ibid; Van Ranst, p. 45, 46; Fleury, cit. n. 51; Hermant, t. 1, c. 48, 51. \\
(35) Nat. Alex. t. 7, c. 3, ar. 6, 9; Van Ranst, p. 47 \& 64; Berti, t. 1, s. 3, c. 3.

\chapter{Heresies of the Fourth Century}

\end{document}